### 2022-09-10 draft.tex
Deploying to gh-pages from @ gottacatchenall/ms_julia_for_ecologists@d199f6d27a498bbadc447c73c51cd36074336e9e 🚀

\begin{itemize}
\tightlist
\item
  Why should ecologists learn julia?

  \begin{itemize}
  \tightlist
  \item
    Well, there are the criteria that are directly measureble that make
    it better than R/Python:

    \begin{itemize}
    \tightlist
    \item
      fast
    \item
      native support on GPUs
    \end{itemize}
  \item
    But there are also criteria that are more subjective, and that take
    experience and practice using the language to appreciate

    \begin{itemize}
    \tightlist
    \item
      clever use of dispatch patterns
    \item
      use of one-lienrs
    \item
      using parameterized types well
    \end{itemize}
  \item
    You will learn how to be a better programmer in \emph{any} language,
    because smart use of julia requires understanding some fundemantal
    concepts in programming that are `hidden' from users in R/python
  \item
    The biggest reason \emph{not} to use julia is that the
    ecology/evolution package ecosystem in R is larger, and the ML
    ecosystem in python is more popular. However:

    \begin{itemize}
    \tightlist
    \item
      you can call \emph{any} R/python function/library using
      RCall/PyCall in julia
    \item
      More packages isn't necessarily better when they don't work
      together
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

In order to measure, understand, and mitigate the consequences of
anthropogenic change on ecosystems the serives they provide, ecologists
need a set of computational tools (Urban \emph{et al.} 2022). These
tools must be performant, but crucially modular and interfaceable
(McIntire \emph{et al.} 2022).

Ecological data is often difficult to access and reuse (Gonzalez \&
Peres-Neto 2015; Poisot \emph{et al.} 2019). Many sources of ecological,
evolutionary, and environmental data exist, but synthesizing this data
into a single product suitable for analysis often remains tedious as
data are not in formats that can be easily combined or interfaced. Here
we propose that we can solve this problem through standardization
(Zimmerman 2008)---developing a common definition such that data
collected in a variety of contexts can be assimilated while minimizing
the overhead of data cleaning and wrangling.

A common representation of ecological data will have three primary
benefits: it will \textbf{1}) enable new forms of analysis by making it
easier to combine data from different sources (Heberling \emph{et al.}
2021), \textbf{2)} enable continuous integration of new data for
next-generation biodiversity monitoring (Kühl \emph{et al.} 2020), and
\textbf{3)} aid in open sharing and reproducability of published results
(Zimmerman 2008; Borregaard \& Hart 2016). Here, we briefly review
approaches to data standardization developed in other fields, in order
to determine what makes an open standard succeed in promoting data
sharing, and what doesn't. Based on the properties of good standards we
identify, we propose building a living standard for ecological data in
the \texttt{Julia} programming language, and argue this is necessary to
obtain the three primary benefits of standardization mentioned earlier.

@@

>>>>

An ecologists guide to learning Julia

\begin{itemize}
\tightlist
\item
  Why should ecologists learn Julia?

  \begin{itemize}
  \tightlist
  \item
    No point ignoring the elephant in the room: R is by far the most
    used language in ecology
  \item
    But many of the problems in contemporary ecology require
    functionality that R just can't handle (mass individual-based
    models, big data management, quick numerical methods)
  \item
    \texttt{julia} has particular features that solve a broad chunk of
    the difficulties of computational analysis of ecological data

    \begin{itemize}
    \tightlist
    \item
      and improves user experience by enabling code that is
      \emph{idiomatic}.
    \end{itemize}
  \item
    this makes julia well suited to become "{[}a platform for ecological
    prediction and forecasting{]}; McIntyre etal, Urban etal
  \end{itemize}
\item
  In what order should people approach new topics

  \begin{itemize}
  \tightlist
  \item
    understand the core concepts that make julia different from other
    languages first

    \begin{itemize}
    \tightlist
    \item
      Types
    \item
      multiple dispatch
    \end{itemize}
  \item
    Learn data management in Julia (emphasize analogies to tidyverse)

    \begin{itemize}
    \tightlist
    \item
      DataFrames
    \end{itemize}
  \item
    Learn stats/ml in Julia (emphasize anologies to GLM in R and
    scikit-learn/tensorflow in python)
  \item
    Learn visualization (Makie)
  \end{itemize}
\item
  Discussion

  \begin{itemize}
  \tightlist
  \item
    Talk about adjacent scientific computing course
  \item
    What does the future of computing in ecology look like?
  \end{itemize}
\end{itemize}

\begin{longtable}[]{@{}l@{}}
\toprule
\endhead
\begin{minipage}[t]{0.05\columnwidth}\raggedright
- Why should ecologists learn julia? - Well, there are the criteria that
are directly measurable that make it better than R/Python: - fast -
native support on GPUs - But there are also criteria that are more
subjective, and that take experience and practice using the language to
appreciate - clever use of dispatch patterns - use of one-lienrs - using
parameterized types well - You will learn how to be a better programmer
in \emph{any} language, because smart use of julia requires
understanding some fundemantal concepts in programming that are `hidden'
from users in R/python - The biggest reason \emph{not} to use julia is
that the ecology/evolution package ecosystem in R is larger, and the ML
ecosystem in python is more popular. However: - you can call \emph{any}
R/python function/library using RCall/PyCall in julia - More packages
isn't necessarily better when they don't work together\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

In order to measure, understand, and mitigate the consequences of
anthropogenic change on ecosystems the services they provide, ecologists
need a modern set of computational tools (Urban \emph{et al.} 2022).

These tools must be performant, but crucially modular and interfaceable
(McIntire \emph{et al.} 2022)

@@

Ecological data is often difficult to access and reuse (Gonzalez \&
Peres-Neto 2015; Poisot \emph{et al.} 2019). Synthesizing data into a
single product suitable for analysis often remains tedious as data are
not in formats that can be easily combined or interfaced.

Here we propose that we can solve this problem through standardization
(Zimmerman 2008)---developing a common definition such that data
collected in a variety of contexts can be assimilated while minimizing
the overhead of data cleaning and wrangling. A common representation of
ecological data will have three primary benefits: it will \textbf{1})
enable new forms of analysis by making it easier to combine data from
different sources (Heberling \emph{et al.} 2021), \textbf{2)} enable
continuous integration of new data for next-generation biodiversity
monitoring (Kühl \emph{et al.} 2020), and \textbf{3)} aid in open
sharing and reproducability of published results (Zimmerman 2008;
Borregaard \& Hart 2016). Here, we briefly review approaches to data
standardization developed in other fields, in order to determine what
makes an open standard succeed in promoting data sharing, and what
doesn't. Based on the properties of good standards we identify, we
propose building a living standard for ecological data in the
\texttt{Julia} programming language, and argue this is necessary to
obtain the three primary benefits of standardization mentioned earlier.
